### 2013-09-17 draft.tex
Started the pakdd paper

\documentclass[english,12pt]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{float}
\usepackage{graphicx}
\usepackage{esint}
\usepackage{natbib}
\usepackage{times,bm}
\usepackage{babel}

\makeatletter
\linespread{1.25}
\pagestyle{headings}

\begin{document}

% \title{}

% \author{}

% \maketitle

\begin{abstract}

\end{abstract}

\section{Introduction}

% \bibliographystyle{apalike}
% \bibliography{financial_market}

\end{document}

### 2013-09-18 draft.tex
Added framework
Added abstract

@@
 
-% \title{}
+\title{Implementation of Parallel Algorithms for ARIMA in Distributed
+  Database Systems}
 
@@
 
-% \maketitle
+\maketitle
 
 \begin{abstract}
-
+  MADlib is an open-source library for scalable in-database
+  analytics. It provides data-parallel implementations of
+  mathematical, statistical and machine-learning methods for
+  structured and unstructured data. We impemented ARIMA in MADlib's
+  framework. The algorithms for fitting ARIMA model to time series are
+  intrinsically sequential, because any calculation for a specific
+  time $t$ depends on the
+  result from the calculation for $t-1$. This makes it difficult to
+  parallelize the ARIMA model fitting. Our solution parallelizes the
+  computation by splitting the data into $n$ chunks. Since the model
+  fitting involves multiple iterations of computations. We use the
+  results from previous iteration as the initial values for each
+  chunk. Thus the computation for each chunk of data does not need to
+  wait for the results from previous chunk. Another technique is used
+  to improve the performance of the implementation, which
+  redistributes the original data so that each chunk can be completely
+  loaded into memory.
 \end{abstract}
@@
 \section{Introduction}
+
+% Introduce to MADlib, and implement ARIMA in MADLib
+
+\section{MADlib}
+
+% More details on MADlib
+
+\section{Implementation of ARIMA}
+
+% A short summary for the next few subsections
+
+\subsection{The Algorithm}
+
+% List the LM algorithm
+
+\subsection{The Problems in Parallelization}
+
+% What are the difficulties that we are facing
+
+\subsection{Our Solution}
+
+% Our solution and how to improve the performance
+
+\subsection{Performance}
+
+% Good performance improvement, some examples
+
+\subsection{Discussion}
+
+% Can be applied to other algorithms
+
+\section{Conclusion}
 

### 2013-09-19 draft.tex
Minor changes to make vim and atp-vim plugin work together
Updates

@@
 
-\begin{abstract}
-  MADlib is an open-source library for scalable in-database
-  analytics. It provides data-parallel implementations of
-  mathematical, statistical and machine-learning methods for
-  structured and unstructured data. We impemented ARIMA in MADlib's
-  framework. The algorithms for fitting ARIMA model to time series are
-  intrinsically sequential, because any calculation for a specific
-  time $t$ depends on the
-  result from the calculation for $t-1$. This makes it difficult to
-  parallelize the ARIMA model fitting. Our solution parallelizes the
-  computation by splitting the data into $n$ chunks. Since the model
-  fitting involves multiple iterations of computations. We use the
-  results from previous iteration as the initial values for each
-  chunk. Thus the computation for each chunk of data does not need to
-  wait for the results from previous chunk. Another technique is used
-  to improve the performance of the implementation, which
-  redistributes the original data so that each chunk can be completely
-  loaded into memory.
+\begin{abstract} MADlib is an open-source library for scalable in-database
+  analytics. We impemented ARIMA in MADlib's framework. The algorithms for
+  fitting ARIMA model to time series are intrinsically sequential, because any
+  calculation for a specific time $t$ depends on the result from the
+  calculation for $t-1$.  This makes it difficult to parallelize the ARIMA
+  model fitting. Our solution parallelizes the computation by splitting the
+  data into $n$ chunks. Since the model fitting involves multiple iterations of
+  computations. We use the results from previous iteration as the initial
+  values for each chunk. Thus the computation for each chunk of data does not
+  need to wait for the results from previous chunk. Another technique is used
+  to improve the performance of the implementation, which redistributes the
+  original data so that each chunk can be completely loaded into memory.
   \end{abstract}
@@
 % Introduce to MADlib, and implement ARIMA in MADLib
+MADlib is an open source library for scalable in-database analytics. It is
+created by the Predictive Analytics Team at Pivotal Inc. MADlib provides
+data-parallel implementations of mathematical, statistical and machine-learning
+methods for structured and unstructured data.
 

### 2013-09-20 draft.tex
Minor changes
Updates

@@
 \usepackage[latin9]{inputenc}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage{esint}
-\usepackage{natbib}
-\usepackage{times,bm}
-\usepackage{babel}
+\usepackage{float} \usepackage{graphicx}
+\usepackage{esint} \usepackage{natbib} \usepackage{times,bm} \usepackage{babel}
+\usepackage{amsmath}
+\usepackage{algorithm2e}
 
@@
 
-\begin{abstract} MADlib is an open-source library for scalable in-database
-  analytics. We impemented ARIMA in MADlib's framework. The algorithms for
-  fitting ARIMA model to time series are intrinsically sequential, because any
-  calculation for a specific time $t$ depends on the result from the
-  calculation for $t-1$.  This makes it difficult to parallelize the ARIMA
-  model fitting. Our solution parallelizes the computation by splitting the
-  data into $n$ chunks. Since the model fitting involves multiple iterations of
+\begin{abstract}
+    MADlib is an open-source library for scalable in-database analytics. We
+    impemented ARIMA in MADlib's framework. The algorithms for fitting ARIMA
+    model to time series are intrinsically sequential, because any calculation
+    for a specific time $t$ depends on the result from the calculation for
+    $t-1$.  This makes it difficult to parallelize the ARIMA model fitting. Our
+    solution parallelizes the computation by splitting the data into $n$
+    chunks. Since the model fitting involves multiple iterations of
     computations. We use the results from previous iteration as the initial
@@
 % Introduce to MADlib, and implement ARIMA in MADLib
-MADlib is an open source library for scalable in-database analytics. It is
-created by the Predictive Analytics Team at Pivotal Inc. MADlib provides
-data-parallel implementations of mathematical, statistical and machine-learning
-methods for structured and unstructured data.
+MADlib \cite{madlib} is an open source library for scalable in-database
+analytics.  It is created by the Predictive Analytics Team at Pivotal Inc.
+MADlib provides data-parallel implementations of mathematical, statistical and
+machine-learning methods for structured and unstructured data.
 
-\section{MADlib}
+\section{What is MADlib?}
 
 % More details on MADlib
+
+MADlib is created by the Predictive Analytics Team at Pivotal Inc. (previously
+Greenplum). It is an add-on package for Greenplum database or PostgreSQL
+database. The package itself is open sourced.
+
+Cohen et al. \cite{mad-skills} explained that "MAD" stands for "magnetic",
+"agile", "deep", and "lib" stands for advanced (mathematical, statistical,
+machine learning), parallel and scalable in-database functions.
+
+The MADlib project was initiated in late 2010. Currently MADlib's latest
+version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
+named PivotalR \cite{pivotalr} is also created by the Predictive Analytics Team
+at Pivotal Inc. A Python wrapper \cite{python-madlib} for MADlib is also
+available.
+
+MADlib can be installed and run on both Grenplum database and PostgreSQL\@. On
+Greenplum database (GPDB) systems, MADlib utilizes the data parallel
+funtionality. The calculation is done on multiple segments of GPDB in parallel,
+and the results from the segments are summarized on the master node. In many
+cases, multiple iterations of such calculations are needed. We use C++ code to
+implement the part inside each iteration, which is computation intensive. We
+use Python code to drive the iterations, which does not have high requirements
+for the performance.
+
+% explain more about the C++ abstraction layer
+
+% more about Python layer
+
+At the time of writing, MADlib has modules for linear regression, logistic
+regression, multinomial logistic regression, elastic-net regularization for
+linear and logistic regressions, all kinds of robust estimators for
+regressions, marginal effects, k-means, association rules, cross validation,
+linear systems, matrix factorization, LDA, data summary, correlation,
+hypothesis tests, SVD and PCA, ARIMA etc and many other supporting functions.
+A very detailed user documentation is available onlin at
+\url{http://madlib.net}.
+
+Here in this paper, we focus our implementation for ARIMA in MADlib.
 
@@
 % List the LM algorithm
+
+An ARIMA model is an auto-regressive integrated moving average model. An ARIMA
+model is typically expressed in the form
+\begin{equation}
+(1 - \phi(B)) Y_t  = (1 + \theta(B)) Z_t,
+\end{equation}
+where $B$ is the backshift operator. The time $t$ is from $1$ to $N$.
+
+ARIMA models involve the following variables:
+\begin{enumerate}
+   \item The lag difference $Y_{t}$, where  $Y_{t} = (1-B)^{d}(X_{t} - \mu)$.
+    \item The values of the time series $X_t$.
+    \item $p$, $q$, and $d$ are the parameters of the ARIMA model.
+      $d$ is the differencing order, $p$ is the order of the AR
+      operator, and $q$ is the order of the MA operator.
+    \item The AR operator $\phi(B)$.
+    \item The MA operator $\theta(B)$.
+    \item The mean value $\mu$, which is always set to be zero for
+      $d>0$ or need to be estimated.
+    \item The error terms $Z_t$.
+\end{enumerate}
+
+The  auto regression operator models the prediction for the next
+observation  as some linear combination of the previous observations.
+More formally, an AR operator of order $p$ is defined as
+\begin{align}
+\phi(B) Y_t= \phi_1 Y_{t-1}   + \dots +  \phi_{p} Y_{t-p}
+\end{align}
+
+The moving average operator is similar, and it models the prediction
+for the next observation as a linear combination of the errors in the
+previous prediction errors.  More formally, the MA operator of order
+$q$ is defined as
+\begin{align}
+\theta(B) Z_t =   \theta_{1} Z_{t-1} + \dots + \theta_{q} Z_{t-q}.
+\end{align}
+
+We assume that
+\begin{equation}
+\Pr(Z_t) = \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}, \quad t > 0
+\end{equation}
+and that  $Z_{-q+1} = Z_{-q+2} = \dots = Z_0 = Z_1 = \dots = Z_p =
+0$. The initial values of $Y_t=X_t-\mu$ for $t=-p+1, -p+2, \dots,
+0$ can be solved from the following linear equations
+\begin{eqnarray}
+\phi_1 Y_0 + \phi_2 Y_{-1} + \cdots + \phi_p Y_{-p+1} &=& Y_1 \nonumber\\
+\phi_2 Y_0 + \cdots + \phi_p Y_{-p+2} &=& Y_2 - \phi_1 Y_1  \nonumber\\
+&\vdots& \nonumber\\
+\phi_{p-1} Y_0 + \phi_p Y_{-1} &=& Y_{p-1} - \phi_1 Y_{p-2} - \cdots -
+\phi_{p-2} Y_1 \nonumber \\
+\phi_p Y_0  &=& Y_p - \phi_1 Y_{p-1} - \cdots - \phi_{p-1} Y_{1} \label{eq:init_Y}
+\end{eqnarray}
+
+The likelihood function $L$ for $N$ values of $Z_t$  is then
+\begin{equation}
+L(\phi, \theta) = \prod_{t = 1}^N  \frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
+\end{equation}
+so the log likelihood function $l$ is
+\begin{align}
+l(\phi, \theta) &= \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
+ \right) \nonumber\\
+ &=  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)  -\frac{Z^2_t}{2 \sigma^2}\nonumber\\
+&=  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
+  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
+\end{align}
+Thus, finding the maximum likelihood is equivalent to solving the
+optimization problem (known as the conditional least squares
+formation)
+\begin{align}
+\min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
+\end{align}
+The error term $Z_t$ can be computed iteratively as follows:
+\begin{align}\label{equ:error-terms}
+Z_t = X_t - F_t(\phi, \theta, \mu)
+\end{align}
+where
+\begin{align}
+F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q \theta_i Z_{t-i}
+\end{align}
+
+In mathematics and computing, the Levenberg-Marquardt algorithm (LMA),
+also known as the damped least-squares (DLS) method, provides a
+numerical solution to the problem of minimizing a function, generally
+nonlinear, over a space of parameters of the function. These
+minimization problems arise especially in least squares curve fitting
+and nonlinear programming.
+
+%Author's (Mark W) Note: This is taken from http://people.duke.edu/~hpgavin/ce281/lm.pdf
+To understand the Levenberg-Marquardt algorithm, it helps to know the
+gradient descent method and the Gauss-Newton method.  On many
+``reasonable'' functions, the gradient descent method takes large
+steps when the current iterate is distant from the true solution, but
+is slow to converge an the current iterate nears the true solution.
+The Gauss-Newton method is much faster for converging when the current
+iterate is in the neighborhood of the true solution.  The
+Levenberg-Marquardt algorithm tries to get the best of best worlds,
+and combine the gradient descent step with Gauss-Newton step in a
+weighted average.  For iterates far from the true solution, the step
+favors the gradient descent step, but as the iterate approaches the
+true solution, the Gauss-Newton step dominates.
+
+%Author's (Mark W) Note: Sudo code taken from http://users.ics.forth.gr/~lourakis/levmar/levmar.pdf
+Like other numeric minimization algorithms, LMA is an iterative
+procedure.  To start a minimization, the user has to provide an
+initial guess for the parameter vector, $p$, as well as some tuning
+parameters $\tau, \epsilon_1, \epsilon_2, \epsilon_3,$ and $k_{max}$.
+Let $Z(p)$ be the vector of calculated errors ($Z_t$'s) for the
+parameter vector $p$, and let $J = (J_{1}, J_{2}, \dots, J_N)^T$
+be a Jacobian matrix.
+
+A proposed implementation is as follows:
+
+\begin{algorithm}[H]
+    \SetAlgoLined
+\alginput{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
+\algoutput{The parameters that maximize the likelihood $\vec{\phi}^*,
+  \vec{\theta}^*, \mu^*$}
+\begin{algorithmic}[1]
+	\State $k \leftarrow 0$  \Comment{Iteration counter}
+	\State $v \leftarrow 2$ \Comment{The change in the weighting factor.}
+	\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$ \Comment{Initialize parameter vector}
+    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
+    equation~\ref{equ:error terms}.  \Comment{Vector of errors}
+	\State $A \leftarrow J^T J$   \Comment{The  Gauss-Newton Hessian approximation}
+	\State $u \leftarrow \tau * \max_i(A_{ii})$ \Comment{Weight of the gradient-descent step}
+	\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \Comment{The gradient descent step.}
+	\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ \Comment{Termination Variable}
+	\While{(not stop) and ($k < k_{max}$)}
+		\State $k \leftarrow k + 1$
+		\Repeat
+			\State $\delta \leftarrow (A + u \times \text{diag}(A))^{-1} g$ \Comment{Calculate step direction}
+			\If{$\| \delta \| \le \epsilon_2 \|
+              (\vec{\phi},\vec{\theta},\mu) \|$} \Comment{Change in the parameters is too small to continue.}
+				\State  $\text{stop} \leftarrow \text{true}$
+			\Else
+				\State $(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new}) \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$ \Comment{Take a trial step in the new direction}
+				\State $\rho \leftarrow (\| Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \| Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2 )/(\delta^T(u \delta + g))$ \Comment{Calculate improvement of trial step}
+				\If{$\rho > 0$} \Comment{Trial step was good, proceed to next iteration}
+					\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$ \Comment{Update variables}
+                    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
+                    equation~\ref{equ:error terms}.
+					\State $A \leftarrow J^T J$
+					\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$
+					\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ or $(\| Z(\vec{\phi},\vec{\theta},\mu)^2 \| \le \epsilon_3)$  \Comment{Terminate if we are close to the solution.}
+					\State $v \leftarrow 2$
+					\State $u \rightarrow u * \max(1/3, 1 - (2\rho - 1)^3 )$
+				\Else  \Comment{Trial step was bad, change weighting on the gradient decent step}
+					\State $v \leftarrow 2 v$
+					\State $u \leftarrow u v$
+				\EndIf
+			\EndIf
+		\Until{(stop) or ($\rho > 0$) }
+	\EndWhile
+	\State $(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$
+\end{algorithmic}
+\label{alg:Levenberg-Marquardt}
+\end{algorithm}
+
+Suggested values for the tuning parameters are $\epsilon_1 = \epsilon_2 =
+\epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{max} = 100$.
 
@@
 
-% \bibliographystyle{apalike}
-% \bibliography{financial_market}
+\bibliographystyle{abbrv}
+
+\bibliography{pred}
 

### 2013-09-22 draft.tex
Minor changes
Added the algorithm
Added more equations in algorithm part

@@
 \usepackage{amsmath}
-\usepackage{algorithm2e}
+%\usepackage{algorithm2e}
+\usepackage[noend]{algpseudocode}
+\usepackage{xspace}
 
@@
 \linespread{1.25}
+
+	% BEGIN Algorithms
+	\theoremstyle{algorithm}
+	\theorembodyfont{\upshape}
+	\newtheorem{algorithm}{Algorithm}[section]
+
+	\newlength{\alglabelwidth}
+	\newcommand{\alginput}[1]{%
+		\par\noindent%
+		\settowidth{\alglabelwidth}{\emph{Output:}}%
+		\makebox[\alglabelwidth][l]{\emph{Input:}} \begin{tabular}[t]{l} #1 \end{tabular}}
+	\newcommand{\algoutput}[1]{%
+		\par\noindent%
+		\settowidth{\alglabelwidth}{\emph{Output:}}%
+		\makebox[\alglabelwidth][l]{\emph{Output:}} \begin{tabular}[t]{l} #1 \end{tabular}}
+	\newcommand{\algprecond}[1]{%
+		\par\noindent\textit{Initialization/Precondition: #1}}
+
+	\newcommand{\set}{\leftarrow}
+	\DeclareMathOperator{\random}{random}
+	\newcommand{\dist}{\ensuremath{\mathit{dist}}}
+	\newcommand{\List}{\mathrm{List}}
+	\newcommand{\Sample}{\mathit{Sample}}
+	\algblockdefx[With]{With}{EndWith}%
+		[1]{\textbf{with} #1 \textbf{do}}%
+		[0]{End}
+	\algnotext[With]{EndWith}
+	% END Algorithms
+
 \pagestyle{headings}
@@
 The error term $Z_t$ can be computed iteratively as follows:
-\begin{align}\label{equ:error-terms}
-Z_t = X_t - F_t(\phi, \theta, \mu)
-\end{align}
+\begin{equation}
+    Z_t = X_t - F_t(\phi, \theta, \mu) \label{eq:error-terms}
+\end{equation}
 where
@@
 
-%Author's (Mark W) Note: This is taken from http://people.duke.edu/~hpgavin/ce281/lm.pdf
 To understand the Levenberg-Marquardt algorithm, it helps to know the
@@
 
-%Author's (Mark W) Note: Sudo code taken from http://users.ics.forth.gr/~lourakis/levmar/levmar.pdf
 Like other numeric minimization algorithms, LMA is an iterative
@@
 
-\begin{algorithm}[H]
-    \SetAlgoLined
+\begin{algorithm}
 \alginput{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
@@
     \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-    equation~\ref{equ:error terms}.  \Comment{Vector of errors}
+    equation~\ref{eq:error-terms}.  \Comment{Vector of errors}
 	\State $A \leftarrow J^T J$   \Comment{The  Gauss-Newton Hessian approximation}
@@
                     \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-                    equation~\ref{equ:error terms}.
+                    equation~\ref{eq:error-terms}.
 					\State $A \leftarrow J^T J$
@@
 \end{algorithmic}
-\label{alg:Levenberg-Marquardt}
+\label{alg:LM}
 \end{algorithm}
@@
 \epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{max} = 100$.
+
+The Jacobian matrix $J = (J_{1}, J_{2}, \dots, J_N)^T$ requires the partial derivatives, which are
+\begin{align}
+J_t = (J_{t, \phi_1}, \dots, J_{t,\phi_p}, J_{t,\theta_1}, \dots,
+J_{t,\theta_q}, J_{t,\mu})^T\ .
+\end{align}
+Here the last term is present only when \texttt{include\_mean} is
+\texttt{True}.
+The iteration relations for $J$ are
+\begin{align}
+J_{t, \phi_i} &= \frac{\partial F_t(\phi,\theta)}{\partial \phi_i} =
+-\frac{\partial Z_t}{\partial \phi_i} = X_{t-i}-\mu + \sum_{j=1}^q
+\theta_j \frac{\partial Z_{t - j}}{\partial \phi_i} = X_{t-i}-\mu - \sum_{j=1}^q
+\theta_j J_{t-j,\phi_i}, \\
+J_{t, \theta_i}&=\frac{\partial F_t(\phi,\theta)}{\partial \theta_i} =
+-\frac{\partial Z_t}{\partial \theta_i} = Z_{t-i} + \sum_{j =1}^q
+\theta_j \frac{\partial Z_{t - j}}{\partial \theta_i} = Z_{t-i} -
+\sum_{j=1}^q \theta_j J_{t-j,\theta_i}, \\
+J_{t, \mu} &=\frac{\partial F_t(\phi,\theta)}{\partial \mu} =
+-\frac{\partial Z_t}{\partial \mu} = 1 -
+\sum_{j=1}^p \phi_j - \sum_{j=1}^q \theta_j \frac{\partial
+  Z_{t-j}}{\partial \mu} = 1 - \sum_{j=1}^p \phi_j - \sum_{j=1}^q
+\theta_j J_{t-j,\mu}.
+\end{align}
+Note that the mean value $\mu$ is considered separately in the above
+formulations. When \texttt{include\_mean} is set to \texttt{False}, $\mu$ will be simply
+set to 0. Otherwise, $\mu$ will also be estimated together with
+$\vec{\phi}$ and $\vec{\theta}$. The initial conditions for the above
+equations are
+\begin{equation}
+J_{t,\phi_i} = J_{t,\theta_j} = J_{t,\mu} = 0 \quad \mbox{for }
+t \leq p, \mbox{ and } i=1,\dots,p; j = 1, \dots, q\ ,
+\end{equation}
+because we have fixed $Z_t$ for $t\leq p$ to be a constant $0$ in the initial
+condition. Note that $J$ is zero not only for $t\leq
+0$ but also for $t\leq p$.
 

### 2013-09-23 draft.tex
Update

--- 2013-09-20 draft.tex
@@
 \documentclass[english,12pt]{article}
+\usepackage{babel}
+\usepackage[latin1]{inputenc}
 \usepackage[T1]{fontenc}
-\usepackage[latin9]{inputenc}
-\usepackage{float} \usepackage{graphicx}
-\usepackage{esint} \usepackage{natbib} \usepackage{times,bm} \usepackage{babel}
 \usepackage{amsmath}
+\usepackage{amssymb}
+
 \usepackage{algorithm2e}
 
-\makeatletter
+%\makeatletter
 \linespread{1.25}
-\pagestyle{headings}
+
+% \pagestyle{headings}
 
@@
 
+%\maketitle
+
 \section{Introduction}
@@
 % Introduce to MADlib, and implement ARIMA in MADLib
-MADlib \cite{madlib} is an open source library for scalable in-database
+MADlib~\cite{madlib} is an open source library for scalable in-database
 analytics.  It is created by the Predictive Analytics Team at Pivotal Inc.
@@
 
-Cohen et al. \cite{mad-skills} explained that "MAD" stands for "magnetic",
-"agile", "deep", and "lib" stands for advanced (mathematical, statistical,
+Cohen et al.~\cite{mad-skills} explained that ``MAD'' stands for ``magnetic'',
+``agile'', ``deep'', and ``lib'' stands for advanced (mathematical, statistical,
 machine learning), parallel and scalable in-database functions.
@@
 version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
-named PivotalR \cite{pivotalr} is also created by the Predictive Analytics Team
-at Pivotal Inc. A Python wrapper \cite{python-madlib} for MADlib is also
+named PivotalR~\cite{pivotalr} is also created by the Predictive Analytics Team
+at Pivotal Inc. A Python wrapper~\cite{python-madlib} for MADlib is also
 available.
@@
 \begin{enumerate}
-   \item The lag difference $Y_{t}$, where  $Y_{t} = (1-B)^{d}(X_{t} - \mu)$.
+   \item The lag difference $Y_{t}$, where  $Y_{t} = {(1-B)}^{d}(X_{t} - \mu)$.
     \item The values of the time series $X_t$.
@@
 More formally, an AR operator of order $p$ is defined as
-\begin{align}
-\phi(B) Y_t= \phi_1 Y_{t-1}   + \dots +  \phi_{p} Y_{t-p}
-\end{align}
+\begin{equation}
+\phi(B) Y_t= \phi_1 Y_{t-1}   + \cdots +  \phi_{p} Y_{t-p}
+\end{equation}
 
@@
 $q$ is defined as
-\begin{align}
-\theta(B) Z_t =   \theta_{1} Z_{t-1} + \dots + \theta_{q} Z_{t-q}.
-\end{align}
+\begin{equation}
+\theta(B) Z_t =   \theta_{1} Z_{t-1} + \cdots + \theta_{q} Z_{t-q}.
+\end{equation}
 
@@
 \end{equation}
-and that  $Z_{-q+1} = Z_{-q+2} = \dots = Z_0 = Z_1 = \dots = Z_p =
+and that  $Z_{-q+1} = Z_{-q+2} = \cdots = Z_0 = Z_1 = \cdots = Z_p =
 0$. The initial values of $Y_t=X_t-\mu$ for $t=-p+1, -p+2, \dots,
@@
 so the log likelihood function $l$ is
-\begin{align}
-l(\phi, \theta) &= \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}} e^{-Z^2_t/2 \sigma^2}
+\begin{eqnarray}
+l(\phi, \theta) &=& \sum_{t = 1}^N \ln \left(\frac{1}{\sqrt{2 \pi \sigma^2}}
+ e^{-Z^2_t/2 \sigma^2}
  \right) \nonumber\\
- &=  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)  -\frac{Z^2_t}{2 \sigma^2}\nonumber\\
-&=  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
-  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
-\end{align}
+ &=&  \sum_{t = 1}^N  - \ln \left( \sqrt{2 \pi \sigma^2}\right)
+-\frac{Z^2_t}{2 \sigma^2}\nonumber\\
+&=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
+  \sigma^2} \sum_{t = 1}^N   Z^2_t\ . \label{eq:loglikelihood}
+\end{eqnarray}
 Thus, finding the maximum likelihood is equivalent to solving the
@@
 formation)
-\begin{align}
+\begin{equation}
 \min_{\theta, \phi} \sum_{t = 1}^N  Z^2_t.
-\end{align}
+\end{equation}
 The error term $Z_t$ can be computed iteratively as follows:
-\begin{align}\label{equ:error-terms}
-Z_t = X_t - F_t(\phi, \theta, \mu)
-\end{align}
+\begin{equation}
+    Z_t = X_t - F_t(\phi, \theta, \mu) \label{eq:error-terms}
+\end{equation}
 where
-\begin{align}
-F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q \theta_i Z_{t-i}
-\end{align}
+\begin{equation}
+F_t(\phi, \theta, \mu) = \mu + \sum_{i=1}^p \phi_i (X_{t-i}-\mu) + \sum_{i=1}^q
+\theta_i Z_{t-i}
+\end{equation}
 
@@
 initial guess for the parameter vector, $p$, as well as some tuning
-parameters $\tau, \epsilon_1, \epsilon_2, \epsilon_3,$ and $k_{max}$.
+parameters $\tau$, $\epsilon_1$, $\epsilon_2$, $\epsilon_3$, and $k_{\max}$.
 Let $Z(p)$ be the vector of calculated errors ($Z_t$'s) for the
-parameter vector $p$, and let $J = (J_{1}, J_{2}, \dots, J_N)^T$
+parameter vector $p$, and let $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$
 be a Jacobian matrix.
@@
 
-\begin{algorithm}[H]
-    \SetAlgoLined
-\alginput{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
-\algoutput{The parameters that maximize the likelihood $\vec{\phi}^*,
+\begin{algorithm}[ht]
+\KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
+\KwOut{The parameters that maximize the likelihood $\vec{\phi}^*,
   \vec{\theta}^*, \mu^*$}
-\begin{algorithmic}[1]
-	\State $k \leftarrow 0$  \Comment{Iteration counter}
-	\State $v \leftarrow 2$ \Comment{The change in the weighting factor.}
-	\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$ \Comment{Initialize parameter vector}
-    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-    equation~\ref{equ:error terms}.  \Comment{Vector of errors}
-	\State $A \leftarrow J^T J$   \Comment{The  Gauss-Newton Hessian approximation}
-	\State $u \leftarrow \tau * \max_i(A_{ii})$ \Comment{Weight of the gradient-descent step}
-	\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \Comment{The gradient descent step.}
-	\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ \Comment{Termination Variable}
-	\While{(not stop) and ($k < k_{max}$)}
-		\State $k \leftarrow k + 1$
-		\Repeat
-			\State $\delta \leftarrow (A + u \times \text{diag}(A))^{-1} g$ \Comment{Calculate step direction}
-			\If{$\| \delta \| \le \epsilon_2 \|
-              (\vec{\phi},\vec{\theta},\mu) \|$} \Comment{Change in the parameters is too small to continue.}
-				\State  $\text{stop} \leftarrow \text{true}$
-			\Else
-				\State $(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new}) \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$ \Comment{Take a trial step in the new direction}
-				\State $\rho \leftarrow (\| Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \| Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2 )/(\delta^T(u \delta + g))$ \Comment{Calculate improvement of trial step}
-				\If{$\rho > 0$} \Comment{Trial step was good, proceed to next iteration}
-					\State $(\vec{\phi},\vec{\theta},\mu) \leftarrow (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$ \Comment{Update variables}
-                    \State Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-                    equation~\ref{equ:error terms}.
-					\State $A \leftarrow J^T J$
-					\State $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$
-					\State $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ or $(\| Z(\vec{\phi},\vec{\theta},\mu)^2 \| \le \epsilon_3)$  \Comment{Terminate if we are close to the solution.}
-					\State $v \leftarrow 2$
-					\State $u \rightarrow u * \max(1/3, 1 - (2\rho - 1)^3 )$
-				\Else  \Comment{Trial step was bad, change weighting on the gradient decent step}
-					\State $v \leftarrow 2 v$
-					\State $u \leftarrow u v$
-				\EndIf
-			\EndIf
-		\Until{(stop) or ($\rho > 0$) }
-	\EndWhile
-	\State $(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$
-\end{algorithmic}
-\label{alg:Levenberg-Marquardt}
+    $k \leftarrow 0$; $v \leftarrow 2$;
+    $(\vec{\phi},\vec{\theta},\mu) \leftarrow
+    (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\;
+    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
+    Eq. (9).  \textit{\footnotesize \# Vector of errors}\;
+    $A \leftarrow J^T J$   \textit{\footnotesize \# The Gauss-Newton Hessian
+    approximation}\;
+    $u \leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize \# Weight of the
+    gradient-descent step}\;
+    $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize \#
+    The gradient descent step}\;
+	$ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$
+    \textit{\footnotesize \# Termination Variable}\;
+	\While{not stop and $k < k_{\max}$}{
+		$k \leftarrow k + 1$\;
+		\Repeat{stop or $\rho > 0$}{
+			$\delta \leftarrow {(A + u \times \text{diag}(A))}^{-1} g$
+            \textit{\footnotesize \# Calculate step direction}\;
+            \eIf(\textit{\footnotesize \# Change  is too small
+              to
+              continue.})
+            {$\| \delta \| \le \epsilon_2 \|
+              (\vec{\phi},\vec{\theta},\mu) \|$} {
+				$\text{stop} \leftarrow \text{true}$\;
+			}{
+				$(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})
+                \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$
+                \textit{\footnotesize \# A trial step}\;
+				$\rho \leftarrow (\|
+                Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \|
+                Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2
+                )/(\delta^T(u \delta + g))$ \;
+                \eIf(\textit{\footnotesize \# Trial step was good}){$\rho >
+                    0$}{
+					$(\vec{\phi},\vec{\theta},\mu) \leftarrow
+                    (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$
+                    \textit{\footnotesize \# Update variables}\;
+                    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
+                    Eq. (9); $A \leftarrow J^T J$; $g \leftarrow J^T
+                    Z(\vec{\phi},\vec{\theta},\mu)$\;
+					$ \text{stop} \leftarrow (\|g\|_{\infty} \le
+                    \epsilon_1)$ or $(\|
+                    Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le
+                    \epsilon_3)$ \;
+                    $v \leftarrow 2$;
+					$u \rightarrow u * \max(1/3, 1 - {(2\rho - 1)}^3 )$\;
+                } (\textit{\footnotesize \# Trial step was bad}){
+                    $v \leftarrow 2 v$;
+					$u \leftarrow u v$\;
+				}
+			}
+		 }
+	}
+	$(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$\;
+
+\label{alg:LM}
 \end{algorithm}
@@
 Suggested values for the tuning parameters are $\epsilon_1 = \epsilon_2 =
-\epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{max} = 100$.
+\epsilon_3 = 10^{-15}, \tau = 10^{-3}$ and $k_{\max} = 100$.
+
+The Jacobian matrix $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$ requires the partial derivatives, which are
+\begin{equation}
+J_t = {(J_{t, \phi_1}, \ldots, J_{t,\phi_p}, J_{t,\theta_1}, \ldots,
+  J_{t,\theta_q}, J_{t,\mu})}^T
+\end{equation}
+Here the last term is present only when \texttt{include\_mean} is
+\texttt{True}.
+The iteration relations for $J$ are
+\begin{eqnarray}
+J_{t, \phi_i} &=& \frac{\partial F_t(\phi,\theta)}{\partial \phi_i} =
+-\frac{\partial Z_t}{\partial \phi_i} = X_{t-i}-\mu + \sum_{j=1}^q
+\theta_j \frac{\partial Z_{t - j}}{\partial \phi_i} = X_{t-i}-\mu - \sum_{j=1}^q
+\theta_j J_{t-j,\phi_i}, \label{eq:J1}\\
+J_{t, \theta_i}&=&\frac{\partial F_t(\phi,\theta)}{\partial \theta_i} =
+-\frac{\partial Z_t}{\partial \theta_i} = Z_{t-i} + \sum_{j =1}^q
+\theta_j \frac{\partial Z_{t - j}}{\partial \theta_i} = Z_{t-i} -
+\sum_{j=1}^q \theta_j J_{t-j,\theta_i}, \label{eq:J2}\\
+J_{t, \mu} &=&\frac{\partial F_t(\phi,\theta)}{\partial \mu} =
+-\frac{\partial Z_t}{\partial \mu} = 1 -
+\sum_{j=1}^p \phi_j - \sum_{j=1}^q \theta_j \frac{\partial
+  Z_{t-j}}{\partial \mu} = 1 - \sum_{j=1}^p \phi_j - \sum_{j=1}^q
+\theta_j J_{t-j,\mu}. \label{eq:J3}
+\end{eqnarray}
+Note that the mean value $\mu$ is considered separately in the above
+formulations. When \texttt{include\_mean} is set to \texttt{False}, $\mu$ will be simply
+set to 0. Otherwise, $\mu$ will also be estimated together with
+$\vec{\phi}$ and $\vec{\theta}$. The initial conditions for the above
+equations are
+\begin{equation}
+J_{t,\phi_i} = J_{t,\theta_j} = J_{t,\mu} = 0 \quad \mbox{for }
+t \leq p,\ \mbox{and }i=1,\dots,p; j = 1, \dots, q\ ,
+\end{equation}
+because we have fixed $Z_t$ for $t\leq p$ to be a constant $0$ in the initial
+condition. Note that $J$ is zero not only for $t\leq
+0$ but also for $t\leq p$.
 
@@
 
-% What are the difficulties that we are facing
+Eqs. (\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
+difficult to parallelize.
 

### 2013-09-24 draft.tex
Made it compilable
Updates

@@
 \usepackage{babel}
+\usepackage{natbib}
 \usepackage[latin1]{inputenc}
@@
 \usepackage{amssymb}
-
 \usepackage{algorithm2e}
@@
 A very detailed user documentation is available onlin at
-\url{http://madlib.net}.
+http://madlib.net.
 
@@
 % A short summary for the next few subsections
+
+In the next few subsections, we will first describe the algorithm that we have
+used to solve the problem of maximization of partial log-likelihood. Then, we
+point out the reason why it is difficult to make the algorithm run parallel.
+The difficulty is not restricted to this specific algorithm. It is a
+common situation in all algorithms that fit ARIMA models. Actually it is easy
+to see that many machine learning models have similar problems. Next, we
+describe our solution to this problem. Our method is quite general and can
+parallelize similar problems. Then we describe a simple method to improve the
+performance of our algorithm. In the end, we discuss the possible ways to
+generalize our methods to more algorithms.
 
@@
 &=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
-  \sigma^2} \sum_{t = 1}^N   Z^2_t\ . \label{eq:loglikelihood}
+  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
 \end{eqnarray}
@@
 
-Eqs. (\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
-difficult to parallelize.
+It is very easy to see that Eqs.
+(\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
+difficult to parallelize. Each step computation uses the result from the
+previous step. Therefore, we have to scan through the data sequentially to
+compute the quatities in these equations.
 

### 2013-09-30 draft.tex
refined intro

@@
 
-% Introduce to MADlib, and implement ARIMA in MADLib
-MADlib~\cite{madlib} is an open source library for scalable in-database
-analytics.  It is created by the Predictive Analytics Team at Pivotal Inc.
-MADlib provides data-parallel implementations of mathematical, statistical and
-machine-learning methods for structured and unstructured data.
+% Introduce to MADlib, and implement ARIMA in MADlib
+MADlib~\cite{madlib} is an open source library for scalable in-database analytics, currently backed by the Predictive Analytics Team at Pivotal Inc.
+It provides data-parallel implementations of mathematical, statistical and machine-learning methods for structured and unstructured data.
 
-\section{What is MADlib?}
+Time series analyses are important in econometrics, mathematical finance, weather forecasting, earthquake prediction and many other fields.
+For example, one of the most conspicuous data analytics task -- stock price forecasting -- falls into the category of time series analysis.
+
+However, different from other common data analytics tasks, time series data have a natural temporal ordering.
+And many time series modeling methods, such as ARIMA~\cite{arima} and Cox proportional hazards~\cite{cox} therefore depend on sequential processing of time series data, which raises a challenge for the data-parallel implementation in MADlib.
+
+In MADlib, we attack this challenge using an idea illustrated below [TODO: add a simple figure to compare MADlib ARIMA and order aggregate with iterations].
+
+\subsection{What is MADlib?}
 
@@
 &=&  -\frac{N}{2} \ln \left( 2 \pi \sigma^2\right)  - \frac{1}{2
-  \sigma^2} \sum_{t = 1}^N   Z^2_t\ \. \label{eq:loglikelihood}
+  \sigma^2} \sum_{t = 1}^N   Z^2_t\ . \label{eq:loglikelihood}
 \end{eqnarray}

### 2013-10-04 draft.tex
Add content to the our solution section

@@
 
+In order to utilize the data-parallel capability of MPP database, we propose to segment the whole time series data into a set of consecutive time series data, numbering from $1$ to $N$, each subset comprising a sequence of consecutive time series data of size $K$.
+Since the model fitting involves multile iterations of computations, for any subset $i$ execpt for the first one, we use the results of the previous subset $i-1$ from the previous iteration as the initial values.
+Thus in each iteration, the computation for each subset of data does not need to wait for the results from the previous subset.
+In this way, the model fitting computaion can be parallelized.
+Besides, we find that aggregating each subset of consecutive time series data into an array (i.e.
+a data chunk) can greatly simpliy the implementation (for example, from a stateful window function implementation to a plain UDF implementation) and accelerate the computation (mainly due to the reduction of I/O overhead and function call overhead).
+Furthermore, we can specify the value of size $K$  according to the system configuration so that each chunk of data can be completely loaded into the avaialble memory).
+
+The following SQL shows how we segment and redistribute the data according to the above proposals.
+
+\begin{verbatim}
+create temp table dist_table as
+  select
+    ((tid - 1) / chunk_size)::integer + 1 as distid, 
+    tid, tval
+  from 
+      input_table
+distributed by (distid)
+\end{verbatim}
+
 \subsection{Performance}

### 2013-10-04 draft.tex
Minor changes
Added figures and the required format style files

--- 2013-09-30 draft.tex
@@
 \documentclass[english,12pt]{article}
+
 \usepackage{babel}
+
 \usepackage{natbib}
+
 \usepackage[latin1]{inputenc}
+
 \usepackage[T1]{fontenc}
+
 \usepackage{amsmath}
+
 \usepackage{amssymb}
+
 \usepackage{algorithm2e}
 
-%\makeatletter
+\usepackage{graphicx}
+
+\makeatletter
 \linespread{1.25}
 
-% \pagestyle{headings}
+\pagestyle{headings}
 
@@
 
+{\bf keywords}: distributiv, parallel computation, database, machine
+learning, big data, ARIMA
+
 %\maketitle
@@
 % Introduce to MADlib, and implement ARIMA in MADlib
-MADlib~\cite{madlib} is an open source library for scalable in-database analytics, currently backed by the Predictive Analytics Team at Pivotal Inc.
-It provides data-parallel implementations of mathematical, statistical and machine-learning methods for structured and unstructured data.
+MADlib~\cite{madlib} is an open source library for scalable
+in-database analytics, currently backed by the Predictive Analytics
+Team at Pivotal Inc. It provides data-parallel implementations of
+mathematical, statistical and machine-learning methods for structured
+and unstructured data.
 
-Time series analyses are important in econometrics, mathematical finance, weather forecasting, earthquake prediction and many other fields.
-For example, one of the most conspicuous data analytics task -- stock price forecasting -- falls into the category of time series analysis.
+Time series analyses are important in econometrics, mathematical
+finance, weather forecasting, earthquake prediction and many other
+fields. For example, one of the most conspicuous data analytics task,
+stock price forecasting, falls into the category of time series
+analysis.
 
-However, different from other common data analytics tasks, time series data have a natural temporal ordering.
-And many time series modeling methods, such as ARIMA~\cite{arima} and Cox proportional hazards~\cite{cox} therefore depend on sequential processing of time series data, which raises a challenge for the data-parallel implementation in MADlib.
+However, different from other common data analytics tasks, time series
+data have a natural temporal ordering. And many time series modeling
+methods, such as ARIMA~\cite{arima} and Cox proportional
+hazards~\cite{cox}, therefore depend on sequential processing of time
+series data, which raises a challenge for the data-parallel
+implementation in MADlib.
 
-In MADlib, we attack this challenge using an idea illustrated below [TODO: add a simple figure to compare MADlib ARIMA and order aggregate with iterations].
+In MADlib, we attack this challenge using an idea illustrated below
+[TODO add a simple figure to compare MADlib ARIMA and order aggregate
+with iterations].
 
@@
 The MADlib project was initiated in late 2010. Currently MADlib's latest
-version is 1.2. By itself MADlib provides a pure SQL interface. An R front-end
-named PivotalR~\cite{pivotalr} is also created by the Predictive Analytics Team
-at Pivotal Inc. A Python wrapper~\cite{python-madlib} for MADlib is also
-available.
+version is 1.3. By itself MADlib provides a pure SQL interface. An R
+front-end named PivotalR~\cite{pivotalr} is also created by the
+Predictive Analytics Team at Pivotal Inc.
 
@@
 
+Although MADlib does not do parallel computation on PostgreSQL, it is
+still valuable for process big data sets that cannot be loaded into
+memory. For example, in this paper we describe our implementation of
+ARIMA in MADlib. We compare our implementation
+
 % explain more about the C++ abstraction layer
@@
 
-At the time of writing, MADlib has modules for linear regression, logistic
-regression, multinomial logistic regression, elastic-net regularization for
-linear and logistic regressions, all kinds of robust estimators for
-regressions, marginal effects, k-means, association rules, cross validation,
-linear systems, matrix factorization, LDA, data summary, correlation,
-hypothesis tests, SVD and PCA, ARIMA etc and many other supporting functions.
-A very detailed user documentation is available onlin at
-http://madlib.net.
+At the time of writing, MADlib has modules for linear regression,
+logistic regression, multinomial logistic regression, elastic-net
+regularization for linear and logistic regressions, all kinds of
+robust estimators for regressions, marginal effects, k-means,
+association rules, cross validation, linear systems, matrix
+factorization, LDA, data summary, correlation, hypothesis tests, SVD
+and PCA, ARIMA etc and many other supporting functions.  A very
+detailed user documentation is available onlin at http://madlib.net.
 
@@
 
-In the next few subsections, we will first describe the algorithm that we have
-used to solve the problem of maximization of partial log-likelihood. Then, we
-point out the reason why it is difficult to make the algorithm run parallel.
-The difficulty is not restricted to this specific algorithm. It is a
-common situation in all algorithms that fit ARIMA models. Actually it is easy
-to see that many machine learning models have similar problems. Next, we
-describe our solution to this problem. Our method is quite general and can
-parallelize similar problems. Then we describe a simple method to improve the
-performance of our algorithm. In the end, we discuss the possible ways to
-generalize our methods to more algorithms.
+In the next few subsections, we will first describe the algorithm that
+we have used to solve the problem of maximization of partial
+log-likelihood. Then, we point out the reason why it is difficult to
+make the algorithm run parallel.  The difficulty is not restricted to
+this specific algorithm. It is a common situation in all algorithms
+that fit ARIMA models. Actually it is easy to see that many machine
+learning models have similar problems. Next, we describe our solution
+to this problem. Our method is quite general and can parallelize
+similar problems. Then we describe a simple method to improve the
+performance of our algorithm. In the end, we discuss the possible ways
+to generalize our methods to other algorithms.
 
@@
 		\Repeat{stop or $\rho > 0$}{
-			$\delta \leftarrow {(A + u \times \text{diag}(A))}^{-1} g$
+            $\delta \leftarrow{(A + u \times \text{diag}(A))}^{-1} g$
             \textit{\footnotesize \# Calculate step direction}\;
@@
 (\ref{eq:error-terms},~\ref{eq:J1},~\ref{eq:J2},~\ref{eq:J3}) are
-difficult to parallelize. Each step computation uses the result from the
-previous step. Therefore, we have to scan through the data sequentially to
-compute the quatities in these equations.
+difficult to parallelize. Each step computation uses the result from
+the previous step. Therefore, we have to scan through the data
+sequentially to compute the quatities in these equations. If the data
+set is very large then it
 
@@
 % Our solution and how to improve the performance
+
+In\ order to utilize the data parallel capability of MPP database, we
+propose to segment the whole time series data into a set of
+consecutive time series data, numbering from $1$ to $N$, each subset
+comprising a sequence of consecutive time series data of size
+$K$. Since the model fitting involves multile iterations of
+computations, for any subset $i$ execpt for the first one, we use the
+results of the $(i-1)$-th subset from the previous iteration as the
+initial values. The first subset's initial values are known to Thus in
+each iteration, the computation for each subset of data does not need
+to wait for the results from the previous subset. In this way, the
+model fitting computaion can be parallelized.
+
+Besides, we find that aggregating each subset of consecutive time
+series data into an array (i.e.\ a data chunk) can greatly simplify
+the implementation (for example, from a stateful window function
+implementation to a plain UDF implementation) and accelerate the
+computation (mainly due to the reduction of I/O overhead and function
+call overhead). Furthermore, we can specify the value of size $K$
+according to the system configuration so that each chunk of data can
+be completely loaded into the avaialble
+memory. Fig. (\ref{fig:seq_vs_dist}) shows the different designs of
+using a sequential implementation and a parallel implementation.
+
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[scale=0.25]{sequential_vs_distributed.png}
+  \end{centering}
+  \caption{\label{fig:seq_vs_dist} The comparison between sequential
+    and distributed designs. }
+\end{figure}
+
+The following SQL shows how we segment and redistribute the data according to the above proposals.
+
+\begin{verbatim}
+create temp table dist_table as
+  select
+    ((tid - 1) / chunk_size)::integer + 1 as distid, 
+    tid, tval
+  from 
+      input_table
+distributed by (distid)
+\end{verbatim}
+
+The chunk size has some effect on the performance, but the effect is
+not very big.
+
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[angle=-90,scale=0.4]{1e8_time_vs_chunksize}
+  \end{centering}
+  \caption{\label{fig:chunksize_effect} We fit a time series with
+    length of $10^8$ using different chunk sizes. The execution times
+    are around $400$ $\sim$ $500$ secs, and are quite stable. }
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[angle=-90,scale=0.4]{gp_time_vs_size}
+  \end{centering}
+  \caption{\label{fig:gp_time_size} We fit a time series with length
+    of $10^8$ using different chunk sizes. The execution times are
+    around $400$ $\sim$ $500$ secs, and are quite stable. }
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[angle=-90,scale=0.4]{pg_time_vs_size}
+  \end{centering}
+  \caption{\label{fig:gp_time_size} We fit a time series with length
+    of $10^8$ using different chunk sizes. The execution times are
+    around $400$ $\sim$ $500$ secs, and are quite stable. }
+\end{figure}
 

### 2013-10-04 draft.tex
use the template
Added a table

@@
-\documentclass[english,12pt]{article}
+\documentclass[english,10pt]{llncs}
 
@@
   \end{centering}
-  \caption{\label{fig:gp_time_size} We fit a time series with length
-    of $10^8$ using different chunk sizes. The execution times are
-    around $400$ $\sim$ $500$ secs, and are quite stable. }
+  \caption{\label{fig:pg_time_size} We fit a time series with length of
+    $10^8$ using different chunk sizes. The execution times are around
+    $400$ $\sim$ $500$ secs, and are quite stable. }
 \end{figure}
+
+\begin{table}[ht]
+  \begin{tabular}{|l|l|l|l|}
+    \hline
+    ~ & MADlib on GPDB & MADlib on Postgres & R's arima function \\ \hline
+    Execution Time (sec) & 364.4 & 1391.9 & 1964.4 \\ \hline
+  \end{tabular}
+  \caption{\label{tbl:compare_R} Here we compare the execution times of ARIMA model fitting in R and in MADlib on Greenplum Database and PostgreSQL database. }
+\end{table}
 

### 2013-10-04 draft.tex
Minor changes

@@
 \begin{algorithm}[ht]
-\KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0, \mu_0$}
-\KwOut{The parameters that maximize the likelihood $\vec{\phi}^*,
-  \vec{\theta}^*, \mu^*$}
-    $k \leftarrow 0$; $v \leftarrow 2$;
-    $(\vec{\phi},\vec{\theta},\mu) \leftarrow
-    (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\;
-    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-    Eq. (9).  \textit{\footnotesize \# Vector of errors}\;
-    $A \leftarrow J^T J$   \textit{\footnotesize \# The Gauss-Newton Hessian
-    approximation}\;
-    $u \leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize \# Weight of the
-    gradient-descent step}\;
-    $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize \#
-    The gradient descent step}\;
-	$ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$
-    \textit{\footnotesize \# Termination Variable}\;
-	\While{not stop and $k < k_{\max}$}{
-		$k \leftarrow k + 1$\;
-		\Repeat{stop or $\rho > 0$}{
-            $\delta \leftarrow{(A + u \times \text{diag}(A))}^{-1} g$
-            \textit{\footnotesize \# Calculate step direction}\;
-            \eIf(\textit{\footnotesize \# Change  is too small
-              to
-              continue.})
-            {$\| \delta \| \le \epsilon_2 \|
-              (\vec{\phi},\vec{\theta},\mu) \|$} {
-				$\text{stop} \leftarrow \text{true}$\;
-			}{
-				$(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})
-                \leftarrow (\vec{\phi},\vec{\theta},\mu) + \delta$
-                \textit{\footnotesize \# A trial step}\;
-				$\rho \leftarrow (\|
+  \KwIn{An initial guess for parameters $\vec{\phi}_0, \vec{\theta}_0,
+    \mu_0$} \KwOut{The parameters that maximize the likelihood
+    $\vec{\phi}^*, \vec{\theta}^*, \mu^*$} $k \leftarrow 0$; $v
+  \leftarrow 2$; $(\vec{\phi},\vec{\theta},\mu) \leftarrow
+  (\vec{\phi}_0,\vec{\theta}_0,\mu_0)$\; Calculate
+  $Z(\vec{\phi},\vec{\theta},\mu)$ with Eq. (9).
+  \textit{\footnotesize // Vector of errors}\; $A \leftarrow J^T J$
+  \textit{\footnotesize // The Gauss-Newton Hessian approximation}\;
+  $u\leftarrow \tau * \max_i(A_{ii})$ \textit{\footnotesize // Weight
+    of the gradient-descent step}\; $g \leftarrow J^T
+  Z(\vec{\phi},\vec{\theta},\mu)$ \textit{\footnotesize // The
+    gradient descent step}\; $ \text{stop} \leftarrow (\|g\|_{\infty}
+  \le \epsilon_1)$ \textit{\footnotesize // Termination Variable}\;
+  \While{not stop and $k < k_{\max}$}{ $k \leftarrow k + 1$\;
+    \Repeat{stop or $\rho > 0$}{ $\delta \leftarrow{(A + u \times
+        \text{diag}(A))}^{-1} g$ \textit{\footnotesize // Calculate
+        step direction}\; \eIf(\textit{\footnotesize // Change is too
+        small to continue.})  {$\| \delta \| \le \epsilon_2 \|
+        (\vec{\phi},\vec{\theta},\mu) \|$} { $\text{stop} \leftarrow
+        \text{true}$\; }{
+        $(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new}) \leftarrow
+        (\vec{\phi},\vec{\theta},\mu) + \delta$ \textit{\footnotesize
+          // A trial step}\; $\rho \leftarrow (\|
         Z(\vec{\phi},\vec{\theta},\mu)\|^2 - \|
         Z(\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})\|^2
-                )/(\delta^T(u \delta + g))$ \;
-                \eIf(\textit{\footnotesize \# Trial step was good}){$\rho >
-                    0$}{
+        )/(\delta^T(u \delta + g))$ \; \eIf(\textit{\footnotesize //
+          Trial step was good}){$\rho > 0$}{
           $(\vec{\phi},\vec{\theta},\mu) \leftarrow
           (\vec{\phi}_{new},\vec{\theta}_{new},\mu_{new})$
-                    \textit{\footnotesize \# Update variables}\;
-                    Calculate $Z(\vec{\phi},\vec{\theta},\mu)$ with
-                    Eq. (9); $A \leftarrow J^T J$; $g \leftarrow J^T
-                    Z(\vec{\phi},\vec{\theta},\mu)$\;
-					$ \text{stop} \leftarrow (\|g\|_{\infty} \le
-                    \epsilon_1)$ or $(\|
-                    Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le
-                    \epsilon_3)$ \;
-                    $v \leftarrow 2$;
-					$u \rightarrow u * \max(1/3, 1 - {(2\rho - 1)}^3 )$\;
-                } (\textit{\footnotesize \# Trial step was bad}){
-                    $v \leftarrow 2 v$;
-					$u \leftarrow u v$\;
-				}
-			}
-		 }
-	}
-	$(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow (\vec{\phi},\vec{\theta},\mu)$\;
+          \textit{\footnotesize // Update variables}\; Calculate
+          $Z(\vec{\phi},\vec{\theta},\mu)$ with Eq. (9); $A \leftarrow
+          J^T J$; $g \leftarrow J^T Z(\vec{\phi},\vec{\theta},\mu)$\;
+          $ \text{stop} \leftarrow (\|g\|_{\infty} \le \epsilon_1)$ or
+          $(\| Z{(\vec{\phi},\vec{\theta},\mu)}^2 \| \le \epsilon_3)$
+          \; $v \leftarrow 2$; $u \rightarrow u * \max(1/3, 1 -
+          {(2\rho - 1)}^3 )$\; } (\textit{\footnotesize // Trial step
+          was bad}){ $v \leftarrow 2 v$; $u \leftarrow u v$\; } } } }
+  $(\vec{\phi}^*,\vec{\theta}^*,\mu^*) \leftarrow
+  (\vec{\phi},\vec{\theta},\mu)$\;
 

### 2013-10-05 draft.tex
Continue

@@
 
+\institute{Predictive Analytics Team, Pivotal Inc.}
+
 \maketitle
@@
 
-{\bf keywords}: distributiv, parallel computation, database, machine
+{\bf keywords}: distributive, parallel computation, database, machine
 learning, big data, ARIMA
@@
   \end{centering}
-  \caption{\label{fig:gp_time_size} We fit a time series with length
-    of $10^8$ using different chunk sizes. The execution times are
-    around $400$ $\sim$ $500$ secs, and are quite stable. }
+  \caption{\label{fig:gp_time_size} We fit time series with different
+    lengths with the chunk size $10^5$ using MADlib's ARIMA function
+    on Greenplum database. The red line is the fit to $t = \alpha l$, where $t$
+    is the execution time and $l$ is the length of the time series.}
 \end{figure}
@@
   \end{centering}
-  \caption{\label{fig:pg_time_size} We fit a time series with length of
-    $10^8$ using different chunk sizes. The execution times are around
-    $400$ $\sim$ $500$ secs, and are quite stable. }
+  \caption{\label{fig:pg_time_size} We fit time series with different
+    lengths with the chunk size $10^5$ using MADlib's ARIMA function
+    on PostgreSQL database. The red line is the fit to $t = \alpha l$, where
+    $t$ is the execution time and $l$ is the length of the time
+    series.}
 \end{figure}
@@
   \end{tabular}
-  \caption{\label{tbl:compare_R} Here we compare the execution times of ARIMA model fitting in R and in MADlib on Greenplum Database and PostgreSQL database. }
+  \caption{\label{tbl:compare_R} Here we compare the execution times
+    of ARIMA model fitting in R and in MADlib on Greenplum Database 
+    and PostgreSQL database. The time series used to fit the ARIMA 
+    model has a length of $10^8$. Note that running MADlib's ARIMA
+    on Postgres is not only faster but also uses much less memory (0.1\%). 
+    Running this data set in R (50G memory) uses almost 70\% of the
+    machine's memory.}
 \end{table}
+
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[angle=-90,scale=0.4]{gp_time_vs_seg}
+  \end{centering}
+  \caption{\label{fig:gp_time_seg} We fit a time series with length of
+    $10^8$ using different numbers of segments in Greenplum
+    database system.}
+\end{figure}
 

### 2013-10-05 draft.tex
Continue

@@
 initial guess for the parameter vector, $p$, as well as some tuning
-parameters $\tau$, $\epsilon_1$, $\epsilon_2$, $\epsilon_3$, and $k_{\max}$.
-Let $Z(p)$ be the vector of calculated errors ($Z_t$'s) for the
-parameter vector $p$, and let $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$
-be a Jacobian matrix.
+parameters $\tau$, $\epsilon_1$, $\epsilon_2$, $\epsilon_3$, and
+$k_{\max}$.  Let $Z(p)$ be the vector of calculated errors ($Z_t$'s)
+for the parameter vector $p$, and let $J = {(J_{1}, J_{2}, \ldots,
+  J_N)}^T$ be a Jacobian matrix.
 
@@
 
-The Jacobian matrix $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$ requires the partial derivatives, which are
+The Jacobian matrix $J = {(J_{1}, J_{2}, \ldots, J_N)}^T$ requires the
+partial derivatives, which are
 \begin{equation}
@@
 Note that the mean value $\mu$ is considered separately in the above
-formulations. When \texttt{include\_mean} is set to \texttt{False}, $\mu$ will be simply
-set to 0. Otherwise, $\mu$ will also be estimated together with
-$\vec{\phi}$ and $\vec{\theta}$. The initial conditions for the above
-equations are
+formulations. When \texttt{include\_mean} is set to \texttt{False},
+$\mu$ will be simply set to 0. Otherwise, $\mu$ will also be estimated
+together with $\vec{\phi}$ and $\vec{\theta}$. The initial conditions
+for the above equations are
 \begin{equation}
@@
 \end{equation}
-because we have fixed $Z_t$ for $t\leq p$ to be a constant $0$ in the initial
-condition. Note that $J$ is zero not only for $t\leq
-0$ but also for $t\leq p$.
+because we have fixed $Z_t$ for $t\leq p$ to be a constant $0$ in the
+initial condition. Note that $J$ is zero not only for $t\leq 0$ but
+also for $t\leq p$.
 
@@
 the previous step. Therefore, we have to scan through the data
-sequentially to compute the quatities in these equations. If the data
-set is very large then it
+sequentially to compute the quatities in these equations, and we have
+to do this in every iteration. If the data set is very large then it
+is time-consuming to use this algorithm.
 
@@
 
-The following SQL shows how we segment and redistribute the data according to the above proposals.
+The following SQL shows how we segment and redistribute the data
+according to the above proposals.
 
@@
     lengths with the chunk size $10^5$ using MADlib's ARIMA function
-    on Greenplum database. The red line is the fit to $t = \alpha l$, where $t$
-    is the execution time and $l$ is the length of the time series.}
+    on Greenplum database. The red line is the fit to $t = \alpha l$,
+    where $t$ is the execution time and $l$ is the length of the time
+    series.}
 \end{figure}
@@
     lengths with the chunk size $10^5$ using MADlib's ARIMA function
-    on PostgreSQL database. The red line is the fit to $t = \alpha l$, where
-    $t$ is the execution time and $l$ is the length of the time
+    on PostgreSQL database. The red line is the fit to $t = \alpha l$,
+    where $t$ is the execution time and $l$ is the length of the time
     series.}
@@
   \caption{\label{fig:gp_time_seg} We fit a time series with length of
-    $10^8$ using different numbers of segments in Greenplum
-    database system.}
+    $10^8$ using different numbers of segments in Greenplum database
+    system.}
 \end{figure}
@@
 \subsection{Performance}
+
+% We may not be able to use Alpine's name here
+
+Some comments about Alpine: Alpine is consistently two to threetimes
+faster than MADlib on GPDB.\@ However, Alpine's result is usually very
+bad if not wrong.
 

### 2013-10-05 draft.tex
intro & references

@@
 
-\title{Implementation of Parallel Algorithms for ARIMA in Distributed
-  Database Systems}
+\title{{Parallel Implementation of Time Series Analysis in Distributed Database Systems}}
 
@@
 
-{\bf keywords}: distributive, parallel computation, database, machine
-learning, big data, ARIMA
+{\bf keywords}: parallel computation, time series, database management system, machine learning, big data, ARIMA
 
@@
 
-% Introduce to MADlib, and implement ARIMA in MADlib
+% Introduce to MADlib
 MADlib~\cite{madlib} is an open source library for scalable
@@
 
+% time series analysis is important
 Time series analyses are important in econometrics, mathematical
@@
 analysis.
-
+% time series analysis is hard to parallelize
 However, different from other common data analytics tasks, time series
@@
 
-In MADlib, we attack this challenge using an idea illustrated below
-[TODO add a simple figure to compare MADlib ARIMA and order aggregate
-with iterations].
+% we focus on ARIMA in this paper
+In this paper, we present our parallel implementation of autoregressive integrated moving average (ARIMA), an popular and important time series analysis model, in MADlib.
+% our idea is chunking
+In MADlib, we attack this mismatch by taking advantage of the iterative nature of the learning algorithms and rely on only local ordering, illustrated in Fig. (\ref{fig:seq_vs_dist}).
+\begin{figure}[ht]
+  \begin{centering}
+    \includegraphics[width=\textwidth]{sequential_vs_distributed.png}
+  \end{centering}
+  \caption{\label{fig:seq_vs_dist} The comparison between sequential
+    and distributed designs. A segment is a running process in parallel databases.}
+\end{figure}
 
@@
 
-\begin{figure}[ht]
-  \begin{centering}
-    \includegraphics[scale=0.25]{sequential_vs_distributed.png}
-  \end{centering}
-  \caption{\label{fig:seq_vs_dist} The comparison between sequential
-    and distributed designs. }
-\end{figure}
-
 The following SQL shows how we segment and redistribute the data

### 2013-10-05 draft.tex
Update the figure 1

@@
 results of the $(i-1)$-th subset from the previous iteration as the
-initial values. The first subset's initial values are known to Thus in
+initial values. The first subset's initial values are known. Thus in
 each iteration, the computation for each subset of data does not need
